---
title: "CYTOF Marker Analysis, final CIR resubmission"
author: "Mari Nakazawa"
date: "2024-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dependencies

```{r dependencies}
library(tidyverse)
library(RColorBrewer)
library(GGally)
library(patchwork)
library(pheatmap)
library(ComplexHeatmap)
library(viridis)
library(ggpubr)
library(MKinfer)
library(broom)
```

## Clinical data

```{r, echo = FALSE}
clinical_cohort <- read_csv("../Data/patients_list_final.csv") %>% 
        select(-`...1`)
```

## Marker expression merged with clinical data

```{r marker expression, echo = FALSE}
markers <- read_csv("../Data/cytof_markers_final.csv")

markers_paired <- markers %>% 
        group_by(patient) %>%
        filter(
    # Ensure each patient has at least one baseline and one on-treatment specimen
                any(timepoint == 0) & any(timepoint %in% c(1, 2))) %>%
        mutate(
    # Determine the earliest on-treatment timepoint for each patient
                EarliestOnTreatment = min(timepoint[timepoint %in% c(1, 2)])) %>%
        filter(
    # Keep rows for baseline (0) and the earliest on-treatment timepoint
                timepoint == 0 | timepoint == EarliestOnTreatment) %>%
        ungroup() %>%
        select(-EarliestOnTreatment) %>%
        mutate(timepoint = recode(timepoint, "0" = "Baseline", "1" = "On Treatment", "2" = "On Treatment")) %>% 
        left_join(clinical_cohort)
```

## Marker data

```{r}
marker_types <- data.frame(Marker = c("CD25", "OX40", "CD28", "CD137", "CCR7", "CD45RA", "CD45RO", "GZMB", "KI67", "TBET", "CTLA4", "LAG3", "PD1", "PDL1", "TIGIT", "TIM3"), Function = c("Activation", "Activation", "Activation", "Activation", "Effector Memory", "Effector Memory", "Effector Memory", "Effector Function/Proliferation", "Effector Function/Proliferation", "Effector Function/Proliferation", "Exhaustion", "Exhaustion", "Exhaustion", "Exhaustion", "Exhaustion", "Exhaustion"))

markers_long <- markers_paired %>% 
        filter(variable %in% marker_types$Marker) %>%
        left_join(marker_types, by = c("variable" = "Marker")) %>% 
        select(patient, timepoint, cluster, variable, Function, value, anti_ctla4, responder) %>% 
        pivot_wider(names_from = timepoint, values_from = value) %>% 
        mutate(log_fc = log2((`On Treatment` + 1e-5)/(Baseline + 1e-5))) 

length(unique(markers_long$patient)) #88
```

## Marker expression by cell type (in MMI)

```{r}
ggplot(markers, aes(x = cluster, y = value)) +
        geom_violin() +
        facet_wrap(~ variable, scales = "free") +
        theme_classic() +
        labs(y = "MMI", x = "Cluster") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Heatmap (logfc) preparation (regimen status)

```{r}
#By cell type
heatmap_data <- markers_long %>% 
        group_by(anti_ctla4, cluster, variable) %>% 
        summarize(average_logfc = mean(log_fc, na.rm = TRUE)) %>% 
        ungroup() 

wide_heatmap <- heatmap_data %>% 
        pivot_wider(names_from = variable, values_from = average_logfc)

scaled_fc <- wide_heatmap %>% 
        select(-cluster, -anti_ctla4) %>% 
        scale() 

col_annotations <- wide_heatmap %>% 
        select(anti_ctla4, cluster)

row_annotations <- marker_types %>% 
        column_to_rownames(var = "Marker")

col_annotations$cluster <- as.factor(col_annotations$cluster)
col_annotations$anti_ctla4 <- factor(col_annotations$anti_ctla4, levels = c(FALSE, TRUE), labels = c("Single ICI", "Combination ICI"))
row_annotations$Function <- factor(row_annotations$Function, levels = c("Activation", "Effector Memory", "Effector Function/Proliferation", "Exhaustion"))
```

## Heatmap using ComplexHeatmap (prep)

```{r}
# Row annotations
cell_function <- unique(row_annotations$Function)
colors_row <- brewer.pal(n = length(cell_function), name = "YlOrRd")

# Create a named vector for the colors
row_colors <- setNames(colors_row, cell_function)

row_anno <- rowAnnotation(
        MarkerFunction = row_annotations$Function,
        col = list(MarkerFunction = row_colors))

# Col annotations
cells <- markers %>% 
        select(cluster) %>% 
        unique() %>% 
        mutate(type = c("B", "B", "DNT", "DNT", "Myeloid", "NK", "NK", "CD8", "CD8", "CD8", "CD8", "CD8", "CD4", "CD4", "CD4", "CD4", "CD4", "CD4", "CD4"))

order <- c("CD25", "OX40", "CD28", "CD137", "CCR7", "CD45RA", "CD45RO", "GZMB", "KI67", "TBET", "CTLA4", "LAG3", "PD1", "PDL1", "TIGIT", "TIM3")

cell_type <- unique(cells$type)
colors_col <- brewer.pal(n = length(cell_type), name = "Set1")

# Create a named vector for the colors
col_colors <- setNames(colors_col, cell_type)

col_anno2 <- HeatmapAnnotation(
        ImmuneCellType = cells$type,
        col = list(ImmuneCellType = col_colors))

col_anno3 <- HeatmapAnnotation(
        MarkerFunction = row_annotations$Function,
        col = list(MarkerFunction = row_colors))
```

## Main heatmap

```{r}
#Split into 2 heatmaps

fc_split <- cbind(col_annotations, scaled_fc) 

fc_split <- fc_split %>%
        split(fc_split$anti_ctla4)

df_combo <- fc_split[["Combination ICI"]]
rownames(df_combo) <- df_combo$cluster
df_combo_matrix <- df_combo %>% 
        select(-cluster, -anti_ctla4) %>% 
        t() %>% 
        .[order, ]

df_single <- fc_split[["Single ICI"]]
rownames(df_single) <- df_single$cluster
df_single_matrix <- df_single %>% 
        select(-cluster, -anti_ctla4) %>% 
        t() %>% 
        .[order, ]

set.seed(123)

ht_df_single <- Heatmap(df_single_matrix, name = "LogFC marker expression", column_title = "Single ICI", top_annotation = col_anno2, right_annotation = row_anno, cluster_rows = FALSE, cluster_columns = FALSE, heatmap_legend_param = list(legend_direction = "horizontal", legend_position = "bottom"))

draw(ht_df_single, heatmap_legend_side = "bottom")

ht_df_combo <- Heatmap(df_combo_matrix, name = "LogFC marker expression", column_title = "Combination ICI", top_annotation = col_anno2, right_annotation = row_anno, cluster_rows = FALSE, cluster_columns = FALSE)

ht_df_single + ht_df_combo 
```

## Function to generate functional marker matrix by cell type

```{r}
#By cell type

order <- c("CD25", "OX40", "CD28", "CD137", "CCR7", "CD45RA", "CD45RO", "GZMB", "KI67", "TBET", "CTLA4", "LAG3", "PD1", "PDL1", "TIGIT", "TIM3")

create_matrix <- function(input_df, cell_cluster) {
        
        d_cell <- input_df %>% 
                filter(cluster == cell_cluster, !is.na(log_fc))
        
        scaled_log_fc <- scale(d_cell$log_fc)
        d_cell <- cbind(d_cell, scaled_log_fc)
        
        matrix_scaled <- d_cell %>% 
                group_by(anti_ctla4, variable) %>%
                summarize(average_logfc = mean(log_fc), .groups = "drop") %>%
                pivot_wider(names_from = variable, values_from = average_logfc) %>%
                select(-anti_ctla4) %>%
                as.data.frame() %>% 
                select(all_of(order)) %>% #reordered
                as.matrix()
        
        rownames(matrix_scaled) <- c("Single ICI", "Combination ICI")
        
        return(matrix_scaled)
}

th17_matrix <- create_matrix(markers_long, "Th17")
treg_matrix <- create_matrix(markers_long, "Treg")
```

## Marker expression in Th17 subset

```{r}
# Mixed effects linear model
library(lme4)
library(emmeans)

# Mixed-effects model
th17_long <- markers_paired %>% 
        filter(cluster == "Th17")

model_th17 <- lmer(value ~ timepoint * anti_ctla4 * variable + (1 | patient), data = th17_long)
summary(model_th17)

# Get estimated marginal means (EMMs)
emms_th17 <- emmeans(model_th17, ~ timepoint * anti_ctla4 | variable)

# Perform pairwise comparisons of the change in immune cell marker expression by single and combination ICI therapy
pairwise_results_th17 <- contrast(emms_th17, interaction = "pairwise", by = "variable")

# View the results
summary(pairwise_results_th17) 

# Test the difference between baseline and on treatment, for single and combination ICI therapy separately
treatment_diff_th17 <- emmeans(model_th17, ~ timepoint | anti_ctla4 * variable)

# Compare changes in expression between treatments
comparison_results_th17 <- contrast(treatment_diff_th17, method = "revpairwise", interaction = TRUE)

# View results
summary(comparison_results_th17)
comparison_df_th17 <- as.data.frame(summary(comparison_results_th17)) %>% 
        mutate(
                sig = case_when(
                        p.value <= 0.001 ~ "***",
                        p.value <= 0.01 ~ "**",
                        p.value <= 0.05 ~ "*",
                        TRUE ~ ""
                )
        )
```

## Annotated heatmap for Th17

```{r}
# Pivot to create a significance matrix
significance_matrix_th17 <- comparison_df_th17 %>%
        filter(timepoint_revpairwise == "On Treatment - Baseline") %>% # Focus on this comparison
        select(variable, anti_ctla4, sig) %>%
        pivot_wider(
                names_from = variable,  # Immune markers become columns
                values_from = sig,  # Significance levels populate the matrix
                names_sort = TRUE
        ) %>% 
        select(all_of(order)) %>% 
        as.matrix()

# Set rownames to anti_ctla4 values
rownames(significance_matrix_th17) <- c("Single ICI", "Combination ICI")

th17_heatmap <- Heatmap(th17_matrix, name = "LogFC marker expression (MMI)", column_title = "Th17 Cell Subset", top_annotation = col_anno3, cell_fun = function(j, i, x, y, width, height, fill) {
                grid.text(significance_matrix_th17[i, j], x, y, gp = gpar(fontsize = 15, col = "black"))},
                row_labels = c("Single ICI", "Combination ICI"), cluster_rows = FALSE, cluster_columns = FALSE)

th17_heatmap
```

## Marker expression in Treg subset

```{r}
# Mixed-effects model
treg_long <- markers_paired %>% 
        filter(cluster == "Treg")

model_treg <- lmer(value ~ timepoint * anti_ctla4 * variable + (1 | patient), data = treg_long)
summary(model_treg)

# Get estimated marginal means (EMMs)
emms_treg <- emmeans(model_treg, ~ timepoint * anti_ctla4 | variable)

# Perform pairwise comparisons of the change in immune cell marker expression by single and combination ICI therapy
pairwise_results_treg <- contrast(emms_treg, interaction = "pairwise", by = "variable")

# View the results
summary(pairwise_results_treg) 

# Test the difference between baseline and on treatment, for single and combination ICI therapy separately
treatment_diff_treg <- emmeans(model_treg, ~ timepoint | anti_ctla4 * variable)

# Compare changes in expression between treatments
comparison_results_treg <- contrast(treatment_diff_treg, method = "revpairwise", interaction = TRUE)

# View results
summary(comparison_results_treg)
comparison_df_treg <- as.data.frame(summary(comparison_results_treg)) %>% 
        mutate(
                sig = case_when(
                        p.value <= 0.001 ~ "***",
                        p.value <= 0.01 ~ "**",
                        p.value <= 0.05 ~ "*",
                        TRUE ~ ""
                )
        )
```

## Annotated heatmap for Treg

```{r}
# Pivot to create a significance matrix
significance_matrix_treg <- comparison_df_treg %>%
        filter(timepoint_revpairwise == "On Treatment - Baseline") %>% # Focus on this comparison
        select(variable, anti_ctla4, sig) %>%
        pivot_wider(
                names_from = variable,  # Immune markers become columns
                values_from = sig,  # Significance levels populate the matrix
                names_sort = TRUE
        ) %>% 
        select(all_of(order)) %>% 
        as.matrix()

# Set rownames to anti_ctla4 values
rownames(significance_matrix_treg) <- c("Single ICI", "Combination ICI")

treg_heatmap <- Heatmap(treg_matrix, name = "LogFC marker expression (MMI)", column_title = "Treg Cell Subset", top_annotation = col_anno3, cell_fun = function(j, i, x, y, width, height, fill) {
                grid.text(significance_matrix_treg[i, j], x, y, gp = gpar(fontsize = 15, col = "black"))},
                row_labels = c("Single ICI", "Combination ICI"), cluster_rows = FALSE, cluster_columns = FALSE)

treg_heatmap
```

# Session info

```{r}
devtools::session_info()
```

This document was processed on: `r Sys.Date()`.